---
title: "HW8"
author: "ASG"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE, cache=TRUE}
knitr::opts_chunk$set(echo = TRUE)
```

## 2

### a
```{r}
library(HRW)  ; library(lattice)  ;  library(nlme)  

# Load in the data:

data(growthIndiana)

# Extract data on males:

growthIndianaMales <- growthIndiana[growthIndiana$male==1,]

# Create relevant variables: 

yOrig <- growthIndianaMales$height
xOrig <- growthIndianaMales$age
idnumOrig <- growthIndianaMales$idnum
typeIsB <- growthIndianaMales$black

# Create new (ordered) ID numbers:

idnum <- rep(NA,length(idnumOrig))
uqID <- unique(idnumOrig)
for (i in 1:length(uqID))
      idnum[idnumOrig==uqID[i]] <- i

# Save mean and standard deviation information:

mean.x <- mean(xOrig)   ;   sd.x <- sd(xOrig)
mean.y <- mean(yOrig)   ;   sd.y <- sd(yOrig)

# Store the standardised versions of the predictor
# and response variables in x and y:

x <- (xOrig - mean.x)/sd.x
y <- (yOrig - mean.y)/sd.y

# Do plot of raw data:

allDataDF <- data.frame(xOrig,yOrig,idnum,typeIsB)
cex.labVal <- 1.35
colourVec <- c("dodgerblue","deeppink","blue","darkmagenta")

figRaw <- xyplot(yOrig~xOrig|idnum,groups=idnum,data=allDataDF,
                 layout=c(9,13),xlab=list(label="age (years)",cex=cex.labVal),
                 ylab=list(label="height (centimetres)",cex=cex.labVal),
                 scales=list(cex=1.25),strip=FALSE,as.table=TRUE,
                 key = list(title="",
                            columns = 2, 
                            points = list(pch = rep(1,2),col = colourVec[1:2]),
                            text = list(c("black","white"),cex=1.55)),
                 panel=function(x,y,subscripts,groups)
                 {
                    panel.grid()
                    colourInd <- 2 - typeIsB[subscripts[1]] 
                    panel.superpose(x,y,subscripts,groups,
                             type="b",col=colourVec[colourInd],pch=1,cex=0.5)

                 })
print(figRaw)

# Set dimension variables:

numObs <- length(x)
numGrp <- length(unique(idnum))

# Set up X and Z matrix for mean curve:

Xbase <- cbind(rep(1,numObs),x)
XBvA <- typeIsB*Xbase
X <- cbind(Xbase,XBvA) 

numIntKnots <- 20
intKnots <- quantile(unique(x),
                     seq(0,1,length=numIntKnots+2))[-c(1,numIntKnots+2)]

range.x <- c(1.01*min(x) - 0.01*max(x),1.01*max(x)-0.01*min(x))

Zbase <- ZOSull(x,intKnots=intKnots,range.x=range.x)
ZBvA <- typeIsB*Zbase

# Set up Z matrix for group specific curves:

numIntKnotsGrp <- 10
intKnotsGrp <- quantile(unique(x),
        seq(0,1,length=numIntKnotsGrp+2))[-c(1,numIntKnotsGrp+2)]

Zgrp <- ZOSull(x,intKnots=intKnotsGrp,range.x=range.x)

# Let up dimension variables:

ncZ <- ncol(Zbase) ; ncZgrp <- ncol(Zgrp)

# Fit using linear mixed model software:

dummyId <- factor(rep(1,numObs))
Zblock <- list(dummyId=pdIdent(~-1+Zbase),
               dummyId=pdIdent(~-1+ZBvA),
               idnum=pdSymm(~x),
               idnum=pdIdent(~-1+Zgrp))
growthINmalGD <- groupedData(y~X[,-1]|rep(1,length=numObs),
                         data=data.frame(y,X,idnum))
cat("Starting fitting (takes several minutes)...\n")
fit <- lme(y~-1+X,data=growthINmalGD,random=Zblock)
cat("Finished fitting.\n")

sig.epsHat <- fit$sigma
sig.uHat <- sig.epsHat*exp(unlist(fit$modelStruct))

lamVal <- (sig.epsHat/sig.uHat)^2

# Set up plotting grids:

ng <- 101
xg <- seq(range.x[1],range.x[2],length=ng)
Xg <- cbind(rep(1,ng),xg)
Zg <- ZOSull(xg,intKnots=intKnots,range.x=range.x)
betaHat <- as.vector(fit$coef$fixed)
uHatBase <- as.vector(fit$coef$random[[1]])
uHatCont <- as.vector(fit$coef$random[[2]])

fhatBaseg <- Xg%*%betaHat[1:2] + Zg%*%uHatBase
Contg <-  Xg%*%betaHat[3:4] + Zg%*%uHatCont
fhatBlackg <- fhatBaseg + Contg

# Plot fitted curves:

Xsubjg <- cbind(rep(1,ng),xg)
Zgrpg <- ZOSull(xg,intKnots=intKnotsGrp,range.x=range.x)
meanCrvs <- vector("list",numGrp)
for (i in 1:numGrp)
{
   indsi <- (1:numObs)[idnum==i]
   uLinHati <- as.vector(fit$coef$random[[3]][i,])
   uSplHati <- as.vector(fit$coef$random[[4]][i,])
   ghati <- Xsubjg%*%uLinHati + Zgrpg%*%uSplHati
   meanCrvs[[i]] <- fhatBaseg + typeIsB[indsi[1]]*Contg + ghati
}

# Convert to original units:

xgOrig <- xgOrig <- mean.x + sd.x*xg
meanCrvsOrig <- vector("list",numGrp)
for (i in 1:numGrp)
   meanCrvsOrig[[i]] <- mean.y + sd.y*meanCrvs[[i]]
   
# Do lattice type plot showing group-specific fits:

figFit <- xyplot(yOrig~xOrig|idnum,groups=idnum,data=allDataDF,
                 layout=c(9,13),xlab=list(label="age (years)",cex=cex.labVal),
                 ylab=list(label="height (centimetres)",cex=cex.labVal),
                 scales=list(cex=1.25),strip=FALSE,as.table=TRUE,
                 key = list(title="",
                            columns = 2, 
                            lines = list(lty = rep(1,2),col = colourVec[3:4]),
                            text = list(c("black","white"),cex=1.55)),
                 panel=function(x,y,subscripts,groups)
                 {
                    panel.grid() 
                    iGrp <- idnum[subscripts][1]
                    colourInd <- 2 - typeIsB[subscripts[1]]
                    panel.superpose(x,y,subscripts,groups,
                             type="b",col=colourVec[colourInd],pch=1,cex=0.5)
                    panel.xyplot(xgOrig,meanCrvsOrig[[iGrp]],
                                 type="l",col=colourVec[colourInd+2])
                 })
print(figFit)

# Obtain penalty matrix:

sigsq.epsHat <- fit$sigma^2
sig.uHat <- as.numeric(sqrt(sigsq.epsHat)*exp(unlist(fit$modelStruct)))

sigsq.Gbl <- sig.uHat[6]^2
sigsq.Gblcont <- sig.uHat[5]^2
cholSigma <- rbind(c(sig.uHat[2],sig.uHat[3]),c(0,sig.uHat[4]))
SigmaHat <- crossprod(cholSigma)
sigsq.Sbj <- sig.uHat[1]^2

DmatGbl <- (sigsq.epsHat/sigsq.Gbl)*diag(ncZ)
DmatGblcont <- (sigsq.epsHat/sigsq.Gblcont)*diag(ncZ)
DmatLinSbj <- sigsq.epsHat*kronecker(diag(numGrp),solve(SigmaHat))
DmatSplSbj <- (sigsq.epsHat/sigsq.Sbj)*diag(ncol(Zgrp)*numGrp)

dimVec <- c(4,nrow(DmatGbl),nrow(DmatGblcont),nrow(DmatLinSbj),
             nrow(DmatSplSbj))

lamMat <- matrix(0,sum(dimVec),sum(dimVec))
csdV <- cumsum(dimVec)
lamMat[(csdV[1]+1):csdV[2],(csdV[1]+1):csdV[2]] <- DmatGbl
lamMat[(csdV[2]+1):csdV[3],
       (csdV[2]+1):csdV[3]] <- DmatGblcont
lamMat[(csdV[3]+1):csdV[4],
       (csdV[3]+1):csdV[4]] <- DmatLinSbj
lamMat[(csdV[4]+1):csdV[5],
       (csdV[4]+1):csdV[5]] <- DmatSplSbj

# Obtain C matrix:

uqID <- unique(idnum)
Cmat <- cbind(X,Zbase,ZBvA)
for (iSbj in 1:numGrp)
{
   newCols <- matrix(0,numObs,2)
   indsCurr <- (1:numObs)[idnum==uqID[iSbj]]
   newCols[indsCurr,] <- X[indsCurr,1:2]
   Cmat <- cbind(Cmat,newCols)
}
for (iSbj in 1:numGrp)
{
   newCols <- matrix(0,numObs,ncol(Zgrp))
   indsCurr <- (1:numObs)[idnum==uqID[iSbj]]
   newCols[indsCurr,] <- Zgrp[indsCurr,]
   Cmat <- cbind(Cmat,newCols)
}

# Obtain full covariance matrix:

CTC <- crossprod(Cmat)
fullCovMat <- solve(CTC + lamMat)

# Find subset of covariance corresponding to the contrast curve:

contInds <- c(3,4,(ncZ+5):(2*ncZ+4))
contCovMat <- fullCovMat[contInds,contInds]

# Obtain approximate pointwise 95% confidence limits:

Cg <- cbind(Xg,Zg)
sdg <- sqrt(sigsq.epsHat)*sqrt(diag(Cg%*%contCovMat%*%t(Cg)))
lowerg <- Contg - qnorm(0.975)*sdg ; upperg <- Contg + qnorm(0.975)*sdg

ContgOrig <- sd.y*Contg
lowergOrig <- sd.y*lowerg
uppergOrig <- sd.y*upperg

# Plot contrast with variability band:

par(mai=c(1,0.9,0.3,0.2))
cex.labVal <- 2   ; cex.axisVal <- 1.5
plot(xgOrig,ContgOrig,bty="l",
     type="n",xlab="age (years)",ylab="mean difference in height (cm)",
     ylim=c(-5,10),cex.lab=cex.labVal,cex.axis=cex.axisVal)

polygon(c(xgOrig,rev(xgOrig)),c(lowergOrig,rev(uppergOrig)),col="PaleGreen",
     border=FALSE)

lines(xgOrig,ContgOrig,col="DarkGreen",lwd=2)
abline(0,0,col="slateblue")
rug(xOrig,col="dodgerblue")
```
the two grid graphs shows the raw grouped data and the other one shows the fitted curve for the same data.

### b

```{r}
compileCode <- TRUE

# Set MCMC sample size paramaters:

nWarm <- 1000
nKept <- 1000
nThin <- 1
   
# Load required packages:

library(HRW)   ;   library(rstan)   ;  library(lattice) 

# Load in the data:

data(growthIndiana)

# Extract data on males:

growthIndianaMales <- growthIndiana[growthIndiana$male == 1,]

# Create relevant variables: 

height <- growthIndianaMales$height
age <- growthIndianaMales$age
idnumOrig <- growthIndianaMales$idnum
typeIsB <- growthIndianaMales$black

# Create new (ordered) ID numbers:

idnum <- rep(NA,length(idnumOrig))
uqID <- unique(idnumOrig)
for (i in 1:length(uqID))
      idnum[idnumOrig == uqID[i]] <- i
  
# Use xOrig and yOrig to denote the original continuous
# predictor and response variables:

xOrig <- age ;  yOrig <- height

# Save mean and standard deviation information:

mean.x <- mean(xOrig)   ;   sd.x <- sd(xOrig)
mean.y <- mean(yOrig)   ;   sd.y <- sd(yOrig)

# Store the standardised versions of the predictor
# and response variables in x and y:

x <- (xOrig - mean.x)/sd.x
y <- (yOrig - mean.y)/sd.y

# Set up X and Z matrix for mean curve:

numObs <- length(x)
numGrp <- length(unique(idnum))

Xbase <- cbind(rep(1,numObs),x)
XB <- typeIsB*Xbase

numIntKnots <- 20
intKnots <- quantile(unique(x),
                     seq(0,1,length=numIntKnots+2))[-c(1,numIntKnots+2)]

range.x <- c(1.01*min(x) - 0.01*max(x),1.01*max(x)-0.01*min(x))
 
Z <- ZOSull(x,intKnots=intKnots,range.x=range.x)
ZA <- (1-typeIsB)*Z
ZB <- typeIsB*Z

# Set up Z matrix for group specific curves:

numIntKnotsGrp <- 10
intKnotsGrp <- quantile(unique(x),
        seq(0,1,length=numIntKnotsGrp+2))[-c(1,numIntKnotsGrp+2)]

Zgrp <- ZOSull(x,intKnots=intKnotsGrp,range.x=range.x)

# Specify hyperparameters:

sigmaBeta <- 1e5  ;  AU <- 1e5
AuGbl <- 1e5      ;  AuGrp <- 1e5
Aeps <- 1e5 

# Let up dimension variables:
   
ncZ <- ncol(Z) ; ncZgrp <- ncol(Zgrp)

# Set up model in Stan:

normGrpSpecContrModel <-
'data
{
   int<lower=1> numObs;             int<lower=1> numGrp;
   int<lower=1> ncZ;                int<lower=1> ncZgrp;
   real<lower=0> sigmaBeta;         real<lower=0> AU;
   real<lower=0> AuGbl;             real<lower=0> AuGrp;
   real<lower=0> Aeps;              matrix[numObs,2] Xbase;          
   matrix[numObs,2] XB;             real y[numObs];                  
   int<lower=1> idnum[numObs];      matrix[numObs,ncZ] ZA;        
   matrix[numObs,ncZ] ZB;           matrix[numObs,ncZgrp] Zgrp;
}
transformed data
{
   vector[2] zeroVec;           zeroVec = rep_vector(0,2);
}
parameters
{
   vector[2] beta;               vector[2] betaBvsA;
   vector[2] U[numGrp];          vector[ncZ] uGblA;
   vector[ncZ] uGblB;            matrix[numGrp,ncZgrp] uGrp;
   cov_matrix[2] SigmaGrp;       real<lower=0> siguGblA;
   real<lower=0> siguGblB;       real<lower=0> siguGrp;
   real<lower=0> sigEps;         vector[2] a;
}
transformed parameters
{
   vector[numObs] meanVec;
   meanVec = Xbase*beta + XB*betaBvsA + ZA*uGblA + ZB*uGblB   
             + to_vector(U[idnum,1]) + to_vector(U[idnum,2]).*col(Xbase,2)
             + rows_dot_product(uGrp[idnum],Zgrp);
}
model
{
   vector[2] scaleSigmaGrp;

   y ~ normal(meanVec,sigEps);

   U ~ multi_normal(zeroVec,SigmaGrp);
   uGblA ~ normal(0,siguGblA);  uGblB ~ normal(0,siguGblB);     
   to_vector(uGrp) ~ normal(0,siguGrp);
      
   a ~ inv_gamma(0.5,pow(AU,-2)); 
   for (k in 1:2) scaleSigmaGrp[k] = 4/a[k];
   SigmaGrp ~ inv_wishart(3,diag_matrix(scaleSigmaGrp));

   beta ~ normal(0,sigmaBeta);   betaBvsA ~ normal(0,sigmaBeta);
   siguGblA ~ cauchy(0,AuGbl);   siguGblB ~ cauchy(0,AuGbl);
   siguGrp ~ cauchy(0,AuGrp);    sigEps ~ cauchy(0,Aeps);
}'

	  
# Set up data list:

allData <- list(numObs=numObs,numGrp=numGrp,ncZ=ncZ,
                ncZgrp=ncZgrp,AU=AU,AuGbl=AuGbl,
                AuGrp=AuGrp,Aeps=Aeps,sigmaBeta=sigmaBeta,
                y=y,idnum=idnum,ZA=ZA,ZB=ZB,
     	        Zgrp=Zgrp,Xbase=Xbase,XB=XB)

# Compile code for model if required:

if (compileCode)
   stanCompilObj <- stan(model_code=normGrpSpecContrModel,data=allData,
                         iter=1,chains=1)

# Perform MCMC:

stanObj <-  stan(model_code=normGrpSpecContrModel,data=allData,warmup=nWarm,
                 iter=(nWarm + nKept),chains=1,thin=nThin,refresh=10,
                 fit=stanCompilObj)

# Save and extract relevant MCMC samples:

betaMCMC <- NULL  ;  betaBvsAMCMC <- NULL   
for (i in 1:2)
{
   charVar1 <- paste("beta[",as.character(i),"]",sep="") 
   betaMCMC <- rbind(betaMCMC,extract(stanObj,charVar1,permuted=FALSE))
   charVar2 <- paste("betaBvsA[",as.character(i),"]",sep="") 
   betaBvsAMCMC <- rbind(betaBvsAMCMC,extract(stanObj,charVar2,permuted=FALSE))
}

uGblAMCMC <- NULL
for (k in 1:ncZ)
{
   charVar <- paste("uGblA[",as.character(k),"]",sep="") 
   uGblAMCMC <- rbind(uGblAMCMC,extract(stanObj,charVar,permuted=FALSE))
}
uGblBMCMC <- NULL
for (k in 1:ncZ)
{
   charVar <- paste("uGblB[",as.character(k),"]",sep="") 
   uGblBMCMC <- rbind(uGblBMCMC,extract(stanObj,charVar,permuted=FALSE))
}
siguGblAMCMC <- as.vector(extract(stanObj,"siguGblA",permuted=FALSE))
siguGblBMCMC <- as.vector(extract(stanObj,"siguGblB",permuted=FALSE))
siguGrpMCMC <- as.vector(extract(stanObj,"siguGrp",permuted=FALSE))
sigEpsMCMC <- as.vector(extract(stanObj,"sigEps",permuted=FALSE))

UMCMC <- NULL
for (i in 1:numGrp)
   for (j in 1:2)
   {
      charVar <- paste("U[",as.character(i),",",as.character(j),"]",sep="")
      UMCMC <- rbind(UMCMC,as.vector(extract(stanObj,charVar,permuted=FALSE)))
   }

uGrpMCMC <- NULL
for (i in 1:numGrp)
   for (j in 1:ncZgrp)
   {
      charVar <- paste("uGrp[",as.character(i),",",as.character(j),"]",sep="")
      uGrpMCMC <- rbind(uGrpMCMC,as.vector(extract(stanObj,charVar,permuted=FALSE)))
   }

# Do lattice-type graphic showing fitted group-specific curves:

AptCol <- "dodgerblue"   ;   AlnCol <- "blue"
BptCol <- "deeppink"     ;   BlnCol <- "maroon"
lwdVal <- 1.5

ng <- 101
xg <- seq(min(x),max(x),length=ng)
Xg <- cbind(rep(1,ng),xg)
Zg <- ZOSull(xg,intKnots=intKnots,range.x=range.x)
Zggrp <- ZOSull(xg,intKnots=intKnotsGrp,range.x=range.x)

fAgMCMC <- Xg%*%betaMCMC + Zg%*%uGblAMCMC
fBgMCMC <- Xg%*%(betaMCMC + betaBvsAMCMC) + Zg%*%uGblBMCMC

meanCrvs <- vector("list",numGrp)
credLows <- vector("list",numGrp)
credUpps <- vector("list",numGrp)
ggMCMC <- vector("list",numGrp)
for (i in 1:numGrp)
{
   # Determine MCMC sample of group-specific deviation
   # grids for group i:

   ggMCMC[[i]] <- (Xg%*%rbind(UMCMC[(2*i-1),],UMCMC[(2*i),])  
                  + Zggrp%*%uGrpMCMC[(i-1)*ncZgrp+1:ncZgrp,])

   # Determine if the current group is Type A or Type B:
  
   indsi <- (1:numObs)[idnum == i]
   currIsB <- typeIsB[indsi[1]]
  
   if (!currIsB)
      currMCMC <-  fAgMCMC + ggMCMC[[i]]

   if (currIsB)
      currMCMC <-  fBgMCMC + ggMCMC[[i]]
   
   meanCrvs[[i]] <- apply(currMCMC,1,mean)
   credLows[[i]] <- apply(currMCMC,1,quantile,0.025)
   credUpps[[i]] <- apply(currMCMC,1,quantile,0.975)
}

# Convert sigEpsMCMC to the original units:

sigEpsMCMCorig <-  sd.y*sigEpsMCMC

# Convert fAgMCMC, fBgMCMC to the original units:

fAgMCMCorig <-  mean.y + sd.y*fAgMCMC
fBgMCMCorig <-  mean.y + sd.y*fBgMCMC

# Convert the group-specific curves to the orginal units:

meanCrvsOrig <- vector("list",numGrp)
credLowsOrig <- vector("list",numGrp)
credUppsOrig <- vector("list",numGrp)

for (i in 1:numGrp)
{
   meanCrvsOrig[[i]] <- mean.y + sd.y*meanCrvs[[i]]
   credLowsOrig[[i]] <- mean.y + sd.y*credLows[[i]]
   credUppsOrig[[i]] <- mean.y + sd.y*credUpps[[i]]
}  
  
# Convert the horizontal plotting grid to the original units:

xgOrig <- mean.x + sd.x*xg

allDataDF <- data.frame(xOrig,yOrig,idnum,typeIsB)

cex.labVal <- 1.35
colourVec <- c("dodgerblue","deeppink","blue","darkmagenta","lightblue","pink")

figFit <- xyplot(yOrig~xOrig|idnum,groups = idnum,data = allDataDF,
                 layout = c(9,13),xlab = list(label = "age (years)",cex = cex.labVal),
                 ylab = list(label = "height (centimetres)",cex = cex.labVal),
                 cex.lab = 4,strip = FALSE,as.table = TRUE,
                 key = list(title = "",
                      columns = 2, 
                      lines = list(lty = rep(1,2),col = colourVec[3:4]),
                      text = list(c("black","white"),cex = 1.55)),
                 panel = function(x,y,subscripts,groups)
                 {
                    panel.grid()
                    iGrp <- panel.number()
                    colourInd <- 2 - typeIsB[subscripts[1]] 
                    panel.superpose(x,y,subscripts,groups,
                                    type = "b",col = colourVec[colourInd],
                                    pch = 1,cex = 0.5)                   
                    panel.polygon(c(xgOrig,rev(xgOrig)),
                                  c(credLowsOrig[[iGrp]],
                                  rev(credUppsOrig[[iGrp]])),
                                  col = colourVec[colourInd+4],border = FALSE)
                    panel.xyplot(xgOrig,meanCrvsOrig[[iGrp]],
                                 col = colourVec[colourInd+2],
                                 type = "l",lwd = lwdVal)
                    
                 })
print(figFit)

# Do contrast function plot:

ContgMCMCorig <- fBgMCMCorig - fAgMCMCorig
ContgMeanOrig <- apply(ContgMCMCorig,1,mean)
ContgLowerOrig <- apply(ContgMCMCorig,1,quantile,0.025)
ContgUpperOrig <- apply(ContgMCMCorig,1,quantile,0.975)

cex.labVal <- 2 ; cex.axisVal <- 1.5
ylimVal <- range(c(ContgLowerOrig,ContgUpperOrig))
par(mai = c(1,0.95,0.1,0.1))
plot(xgOrig,ContgMeanOrig,bty = "l",type = "n",xlab = "age (years)",
     ylab = "mean difference in height (centimetres)",
     ylim = ylimVal,cex.lab = cex.labVal,cex.axis = cex.axisVal)

polygon(c(xgOrig,rev(xgOrig)),c(ContgLowerOrig,rev(ContgUpperOrig)),
         col = "palegreen",border = FALSE)
lines(xgOrig,ContgMeanOrig,col = "darkgreen",lwd = 2)
abline(0,0,col = "slateblue")
nSub <- 1000
subInds <- sample(1:length(x),nSub,replace = FALSE)
rug(xOrig[subInds],col = "dodgerblue")

# Do MCMC summary plot to check convergence:

indQ1 <- length(xgOrig[xgOrig<quantile(xOrig,0.25)])
indQ2 <- length(xgOrig[xgOrig<quantile(xOrig,0.50)])
indQ3 <- length(xgOrig[xgOrig<quantile(xOrig,0.75)])
ContQ1MCMCorig <- ContgMCMCorig[indQ1,]
ContQ2MCMCorig <- ContgMCMCorig[indQ2,]
ContQ3MCMCorig <- ContgMCMCorig[indQ3,]

MCMClist <- list(cbind(sigEpsMCMCorig,ContQ1MCMCorig,ContQ2MCMCorig,ContQ3MCMCorig))
parNamesVal <- list(expression(sigma[epsilon]),
                    c("contrast function","at 1st quantile","of age"),
                    c("contrast function","at 2nd quantile","of age"),
                    c("contrast function","at 3rd quantile","of age"))
summMCMC(MCMClist,parNames = parNamesVal)

############ End of maleGrowthIndianaBayes ###########
```

In the diagnostic plots , we can see that the trace behaves normally and is as usual. We can also see in the lag scatterplot , we have a linear relationship which might be a bit concerning for all the distributions. Transforming the variables might fix it ? I am not really sure

### c


### 4.1

### a
```{r}
########## R script: femSBMDbayes ##########

# For conducting a Bayesian additive mixed
# model analysis on the spinal bone mineral
# density data using Markov chain Monte Carlo 
# and Stan.

# Last changed: 29 AUG 2018

# Set flag for code compilation (needed if 
# running script first time in current session):

compileCode <- TRUE

# Set MCMC sample size paramaters:

nWarm <- 1000
nKept <- 1000
nThin <- 1

# Load required packages:

library(HRW)   ;   library(rstan)   ;   library(lattice)

# Load in the female spinal bone data and extract the 
# component variables:

data(femSBMD)
femSBMD$asian<-ifelse(femSBMD$ethnicity=='Asian',1,0)
idnum <- femSBMD$idnum 
x1 <- femSBMD$black 
x2 <- femSBMD$hispanic 
x3 <- femSBMD$white
x0 <- femSBMD$asian
x4Orig <- femSBMD$age 
yOrig <- femSBMD$spnbmd

# Standardise data for Bayesian analysis:

mean.x4 <- mean(x4Orig) ; sd.x4 <- sd(x4Orig)
mean.y <- mean(yOrig)   ; sd.y <- sd(yOrig)
x4 <- (x4Orig - mean.x4)/sd.x4
y <- (yOrig - mean.y)/sd.y

# Set up matrices for additive model:

numObs <- length(y)
X <- cbind(rep(1,numObs),x0,x1,x2,x3,x4)
ncX <- ncol(X)

numIntKnots <- 15
intKnots <- quantile(unique(x4),seq(0,1,length = 
                 (numIntKnots+2))[-c(1,(numIntKnots+2))])
range.x4 <- c(1.01*min(x4)-0.01*max(x4),1.01*max(x4)-0.01*min(x4))
Zspl <- ZOSull(x4,intKnots = intKnots,range.x = range.x4)
ncZ <- ncol(Zspl)
Zspl_asian<-X[,2]*Zspl;Zspl_black<-X[,3]*Zspl;
Zspl_hispanic<-X[,4]*Zspl;Zspl_white<-X[,5]*Zspl;

numGrp <- length(unique(idnum))
numObs <- length(y)

# Set hyperparameters:

sigmaBeta <- 1e5 ; AU <- 1e5 ; Au_asian <- 1e5 ; Au_black <- 1e5 ; Au_hispanic <- 1e5 ; Au_white <- 1e5 ; Aeps <- 1e5

# Specify model in Stan:

addMixModModel <- 
'data
{
   int<lower=1> numObs;         int<lower=1> numGrp;
   int<lower=1> ncX;            int<lower=1> ncZ;        
   real<lower=0> sigmaBeta;     real<lower=0> AU;
   real<lower=0> Aeps;          real<lower=0> Au_asian;
   vector[numObs] y;            int<lower=1> idnum[numObs];
   matrix[numObs,ncX] X;        matrix[numObs,ncZ] Zspl_asian;
   real<lower=0> Au_black;           real<lower=0> Au_hispanic;
   real<lower=0> Au_white;      matrix[numObs,ncZ] Zspl_black;
   matrix[numObs,ncZ] Zspl_hispanic;  matrix[numObs,ncZ] Zspl_white;
}
parameters
{
   vector[ncX] beta;            vector[numGrp] U;            
   vector[ncZ] u_asian;         real<lower=0> sigmaU;        
   real<lower=0> sigmau_asian;        real<lower=0> sigmaEps;    
   vector[ncZ] u_black;         vector[ncZ] u_hispanic;
   vector[ncZ] u_white;         real<lower=0> sigmau_black;
   real<lower=0> sigmau_hispanic;        real<lower=0> sigmau_white; 
}
model
{
   y ~ normal(X*beta + U[idnum] + Zspl_asian*u_asian+Zspl_black*u_black+Zspl_hispanic*u_hispanic+Zspl_white*u_white,sigmaEps);
   U ~ normal(0,sigmaU);           u_asian ~ normal(0,sigmau_asian);
   beta  ~ normal(0,sigmaBeta) ;   sigmaEps ~ cauchy(0,Aeps);
   sigmaU ~ cauchy(0,AU) ;   sigmau_asian ~ cauchy(0,Au_asian);
   u_black ~ normal(0,sigmau_black);  sigmau_black ~ cauchy(0,Au_black);
   u_hispanic ~ normal(0,sigmau_hispanic);  sigmau_hispanic ~ cauchy(0,Au_hispanic);
   u_white ~ normal(0,sigmau_white);  sigmau_white ~ cauchy(0,Au_white);
}'
# Zspl*u_black*X[,3]+Zspl*u_hispanic*X[,3]+Zspl*u_white*X[,4]
# Fit model using MCMC via Stan:

allData <- list(numObs = numObs,numGrp = numGrp,ncX = ncX,ncZ = ncZ,
                idnum = idnum,X = X,y = y,Zspl_asian = Zspl_asian,sigmaBeta = sigmaBeta,
                AU = AU,Au_asian = Au_asian,Au_black = Au_black,Au_hispanic = Au_hispanic,
                Au_white = Au_white,Aeps = Aeps,Zspl_black = Zspl_black,
                Zspl_hispanic = Zspl_hispanic,Zspl_white = Zspl_white)

# Compile code for model if required:

if (compileCode)
   stanCompilObj <- stan(model_code = addMixModModel,data = allData,
                         iter = 1,chains = 1)

# Perform MCMC:

stanObj <-  stan(model_code = addMixModModel,data = allData,warmup = nWarm,
                 iter = (nWarm + nKept),chains = 1,thin = nThin,refresh = 100,
                 fit = stanCompilObj)

# Save and extract relevant MCMC samples:

beta0MCMC <- as.vector(extract(stanObj,"beta[1]",permuted = FALSE))
beta1MCMC <- as.vector(extract(stanObj,"beta[2]",permuted = FALSE))
beta2MCMC <- as.vector(extract(stanObj,"beta[3]",permuted = FALSE))
beta3MCMC <- as.vector(extract(stanObj,"beta[4]",permuted = FALSE))
beta4MCMC <- as.vector(extract(stanObj,"beta[5]",permuted = FALSE))

sigmaUMCMC <- as.vector(extract(stanObj,"sigmaU",permuted = FALSE))
sigmaEpsMCMC <- as.vector(extract(stanObj,"sigmaEps",permuted = FALSE))


UMCMC <- NULL
for (k in 1:numGrp)
{
   charVar <- paste("U[",as.character(k),"]",sep = "") 
   UMCMC <- rbind(UMCMC,extract(stanObj,charVar,permuted = FALSE))
}

uMCMC_asian <- NULL
for (k in 1:ncZ)
{
   charVar <- paste("u_asian[",as.character(k),"]",sep = "") 
   uMCMC_asian <- rbind(uMCMC_asian,extract(stanObj,charVar,permuted = FALSE))
}

uMCMC_black <- NULL
for (k in 1:ncZ)
{
  charVar <- paste("u_black[",as.character(k),"]",sep = "") 
  uMCMC_black <- rbind(uMCMC_black,extract(stanObj,charVar,permuted = FALSE))
}

uMCMC_hispanic <- NULL
for (k in 1:ncZ)
{
  charVar <- paste("u_hispanic[",as.character(k),"]",sep = "") 
  uMCMC_hispanic <- rbind(uMCMC_hispanic,extract(stanObj,charVar,permuted = FALSE))
}

uMCMC_white <- NULL
for (k in 1:ncZ)
{
  charVar <- paste("u_white[",as.character(k),"]",sep = "") 
  uMCMC_white <- rbind(uMCMC_white,extract(stanObj,charVar,permuted = FALSE))
}

# Convert to parameters original scale:

beta1MCMCorig <- beta1MCMC*sd.y
beta2MCMCorig <- beta2MCMC*sd.y
beta3MCMCorig <- beta3MCMC*sd.y
beta4MCMCorig <- beta4MCMC*(sd.y/sd.x4)

beta0MCMCorig <-  mean.y + sd.y*beta0MCMC -  mean.x4*beta4MCMCorig

sigmaUMCMCorig <- sigmaUMCMC*sd.y
sigmaEpsMCMCorig <- sigmaEpsMCMC*sd.y

# Do parameters plot:

parms <- list(cbind(beta1MCMCorig,beta2MCMCorig,beta3MCMCorig,
                    sigmaUMCMCorig,sigmaEpsMCMCorig))
parNamesVal <- list(c("Black","vs. Asian"),c("Hispanic","vs. Asian"),
                      c("White","vs. Asian"),c(expression(sigma[U])),
                      c(expression(sigma[epsilon])))
summMCMC(parms,parNames = parNamesVal)

# Do fitted curves using lattice graphics:

ng <- 101
ylim.val <- c(min(yOrig),max(yOrig))
x4g <- seq(min(x4),max(x4),length = ng)
Xg <- cbind(rep(1,ng),x4g)
Zg <- ZOSull(x4g,intKnots = intKnots,range.x = range.x4)

midCurvsg <- list()
lowCurvsg <- list()
uppCurvsg <- list()

for (ip in 1:4)
{
   if (ip == 1){
     betaSplMCMC <- rbind(beta0MCMC,beta1MCMC)

     fMCMC <-   Xg%*%betaSplMCMC + Zg%*%uMCMC_asian
   } 
   if (ip == 2){
     betaSplMCMC <- rbind(beta0MCMC+beta2MCMC,beta1MCMC)
     fMCMC <-   Xg%*%betaSplMCMC + Zg%*%uMCMC_black
     
   } 
   if (ip == 3){
     betaSplMCMC <- rbind(beta0MCMC+beta3MCMC,beta1MCMC)

     fMCMC <-   Xg%*%betaSplMCMC + Zg%*%uMCMC_hispanic
     
   }
   if (ip == 4){
     betaSplMCMC <- rbind(beta0MCMC+beta4MCMC,beta1MCMC)
     fMCMC <-   Xg%*%betaSplMCMC + Zg%*%uMCMC_white
     
   }

   
   fMCMCorig <- fMCMC*sd.y + mean.y
   credLowerOrig <- apply(fMCMCorig,1,quantile,0.025)
   credUpperOrig <- apply(fMCMCorig,1,quantile,0.975)
   fhatgOrig <- apply(fMCMCorig,1,mean)

   x4gOrig <- x4g*sd.x4 + mean.x4

   midCurvsg[[ip]] <- apply(fMCMCorig,1,mean)
   lowCurvsg[[ip]] <- apply(fMCMCorig,1,quantile,0.025)
   uppCurvsg[[ip]] <- apply(fMCMCorig,1,quantile,0.975)
}

fitFig <- xyplot(yOrig~x4Orig|ethnicity,groups = idnum,
                  as.table = TRUE,data = femSBMD,
                  strip = strip.custom(par.strip.text = list(cex = 1.5)),
                  par.settings = list(layout.heights = list(strip = 1.6)),
                  scales = list(cex = 1.25),
                  xlab = list("age (years)",cex = 1.5),
                  ylab = list(expression(paste(
                  "spinal bone mineral density (g/c",m^2,")")),cex = 1.5),
                  subscripts = TRUE,
                  panel = function(x,y,subscripts,groups)
                  {
                     panel.grid() 
                     if (any(femSBMD$ethnicity[subscripts] == "Asian"))    panNum <- 1
                     if (any(femSBMD$ethnicity[subscripts] == "Black"))    panNum <- 2
                     if (any(femSBMD$ethnicity[subscripts] == "Hispanic")) panNum <- 3
                     if (any(femSBMD$ethnicity[subscripts] == "White"))    panNum <- 4

                     panel.superpose(x,y,subscripts,groups,type = "b",pch = 16)

                     panel.polygon(c(x4gOrig,rev(x4gOrig)),c(lowCurvsg[[panNum]],rev(uppCurvsg[[panNum]])),
                                   col = "palegreen",border = FALSE)

                     panel.xyplot(x4gOrig,midCurvsg[[panNum]],lwd = 2,type = "l",col = "darkgreen")     
                })
   
print(fitFig)

########## End of femSBMDbayes ##########
```

## Black-Asian Contrast Curve

```{r}
# Plot contrast curve fit:

estFunCol <- "darkgreen"; 
varBandCol <- "palegreen"

betaSplMCMC <- rbind(beta0MCMC,beta1MCMC)
ContrastMCMC1 <- Xg%*%betaSplMCMC + Zg%*%(uMCMC_black-uMCMC_asian)
credLower1 <- apply(ContrastMCMC1,1,quantile,0.025)
credUpper1 <- apply(ContrastMCMC1,1,quantile,0.975)
Contrastg1 <- apply(ContrastMCMC1,1,mean)

# Convert to original units:

ContrastgOrig1 <- sd.y*Contrastg1
credLowerOrig1 <- sd.y*credLower1 
credUpperOrig1<- sd.y*credUpper1 

par(mfrow = c(1,1),mai = c(1.02,0.9,0.82,0.42))
plot(0,0,type = "n",bty = "l",xlim = range(xgOrig),
     range(c(credLowerOrig1,credUpperOrig1)),
     main="Black-Asian Contrast Curve Contrast Curve",
     xlab = "Age",
     ylab = " Spinal Bone Mineral Density",
     cex.lab = cex.labVal,cex.axis = cex.axisVal)

polygon(c(xgOrig,rev(xgOrig)),c(credLowerOrig1,rev(credUpperOrig1)),
        col = varBandCol,border = FALSE)

lines(xgOrig,ContrastgOrig1,lwd = 2,col = estFunCol)
abline(0,0,col = "slateblue",lwd = 2)
```

## Hispanic-Asian Contrast Curve

```{r}
# Plot contrast curve fit:

estFunCol <- "darkgreen"; 
varBandCol <- "palegreen"
betaSplMCMC <- rbind(beta0MCMC,beta1MCMC)
ContrastMCMC2 <- Xg%*%betaSplMCMC + Zg%*%(uMCMC_hispanic-uMCMC_asian)

credLower2 <- apply(ContrastMCMC2,1,quantile,0.025)
credUpper2 <- apply(ContrastMCMC2,1,quantile,0.975)
Contrastg2 <- apply(ContrastMCMC2,1,mean)

# Convert to original units:

ContrastgOrig2 <- sd.y*Contrastg2
credLowerOrig2 <- sd.y*credLower2 
credUpperOrig2<- sd.y*credUpper2

par(mfrow = c(1,1),mai = c(1.02,0.9,0.82,0.42))
plot(0,0,type = "n",bty = "l",xlim = range(xgOrig),
     range(c(credLowerOrig2,credUpperOrig2)),
     main="Hispanic-Asian Contrast Curve Contrast Curve",
     xlab = "Age",
     ylab = " Spinal Bone Mineral Density",
     cex.lab = cex.labVal,cex.axis = cex.axisVal)

polygon(c(xgOrig,rev(xgOrig)),c(credLowerOrig2,rev(credUpperOrig2)),
        col = varBandCol,border = FALSE)

lines(xgOrig,ContrastgOrig2,lwd = 2,col = estFunCol)
abline(0,0,col = "slateblue",lwd = 2)
```

## White-Asian Contrast Curve

```{r}
# Plot contrast curve fit:

estFunCol <- "darkgreen"; 
varBandCol <- "palegreen"
betaSplMCMC <- rbind(beta0MCMC,beta1MCMC)
ContrastMCMC3 <- Xg%*%betaSplMCMC + Zg%*%(uMCMC_white-uMCMC_asian)

credLower3 <- apply(ContrastMCMC3,1,quantile,0.025)
credUpper3 <- apply(ContrastMCMC3,1,quantile,0.975)
Contrastg3 <- apply(ContrastMCMC3,1,mean)

# Convert to original units:

ContrastgOrig3 <- sd.y*Contrastg3
credLowerOrig3 <- sd.y*credLower3 
credUpperOrig3<- sd.y*credUpper3

par(mfrow = c(1,1),mai = c(1.02,0.9,0.82,0.42))
plot(0,0,type = "n",bty = "l",xlim = range(xgOrig),
     range(c(credLowerOrig3,credUpperOrig3)),
     main="White-Asian Contrast Curve Contrast Curve",
     xlab = "Age",
     ylab = " Spinal Bone Mineral Density",
     cex.lab = cex.labVal,cex.axis = cex.axisVal)

polygon(c(xgOrig,rev(xgOrig)),c(credLowerOrig3,rev(credUpperOrig3)),
        col = varBandCol,border = FALSE)

lines(xgOrig,ContrastgOrig3,lwd = 2,col = estFunCol)
abline(0,0,col = "slateblue",lwd = 2)
```

## Black-Hispanic Contrast Curve

```{r}
# Plot contrast curve fit:

estFunCol <- "darkgreen"; 
varBandCol <- "palegreen"
betaSplMCMC <- rbind(beta0MCMC,beta1MCMC)
ContrastMCMC4 <- Xg%*%betaSplMCMC + Zg%*%(uMCMC_black-uMCMC_hispanic)

credLower4 <- apply(ContrastMCMC4,1,quantile,0.025)
credUpper4 <- apply(ContrastMCMC4,1,quantile,0.975)
Contrastg4 <- apply(ContrastMCMC4,1,mean)

# Convert to original units:

ContrastgOrig4 <- sd.y*Contrastg4
credLowerOrig4 <- sd.y*credLower4 
credUpperOrig4<- sd.y*credUpper4

par(mfrow = c(1,1),mai = c(1.02,0.9,0.82,0.42))
plot(0,0,type = "n",bty = "l",xlim = range(xgOrig),
     range(c(credLowerOrig4,credUpperOrig4)),
     main="Black-Hispanic Contrast Curve Contrast Curve",
     xlab = "Age",
     ylab = " Spinal Bone Mineral Density",
     cex.lab = cex.labVal,cex.axis = cex.axisVal)

polygon(c(xgOrig,rev(xgOrig)),c(credLowerOrig4,rev(credUpperOrig4)),
        col = varBandCol,border = FALSE)

lines(xgOrig,ContrastgOrig4,lwd = 2,col = estFunCol)
abline(0,0,col = "slateblue",lwd = 2)
```

## White-Hispanic Contrast Curve

```{r}
# Plot contrast curve fit:

estFunCol <- "darkgreen"; 
varBandCol <- "palegreen"
betaSplMCMC <- rbind(beta0MCMC,beta1MCMC)
ContrastMCMC5 <- Xg%*%betaSplMCMC + Zg%*%(uMCMC_white-uMCMC_hispanic)

credLower5 <- apply(ContrastMCMC5,1,quantile,0.025)
credUpper5 <- apply(ContrastMCMC5,1,quantile,0.975)
Contrastg5 <- apply(ContrastMCMC5,1,mean)

# Convert to original units:

ContrastgOrig5 <- sd.y*Contrastg5
credLowerOrig5 <- sd.y*credLower5 
credUpperOrig5<- sd.y*credUpper5

par(mfrow = c(1,1),mai = c(1.02,0.9,0.82,0.42))
plot(0,0,type = "n",bty = "l",xlim = range(xgOrig),
     range(c(credLowerOrig5,credUpperOrig5)),
     main="White-Hispanic Contrast Curve Contrast Curve",
     xlab = "Age",
     ylab = " Spinal Bone Mineral Density",
     cex.lab = cex.labVal,cex.axis = cex.axisVal)

polygon(c(xgOrig,rev(xgOrig)),c(credLowerOrig5,rev(credUpperOrig5)),
        col = varBandCol,border = FALSE)

lines(xgOrig,ContrastgOrig5,lwd = 2,col = estFunCol)
abline(0,0,col = "slateblue",lwd = 2)
```

## White-Black Contrast Curve

```{r}
# Plot contrast curve fit:

estFunCol <- "darkgreen"; 
varBandCol <- "palegreen"
betaSplMCMC <- rbind(beta0MCMC,beta1MCMC)
ContrastMCMC6 <- Xg%*%betaSplMCMC + Zg%*%(uMCMC_white-uMCMC_black)

credLower6 <- apply(ContrastMCMC6,1,quantile,0.025)
credUpper6 <- apply(ContrastMCMC6,1,quantile,0.975)
Contrastg6 <- apply(ContrastMCMC6,1,mean)

# Convert to original units:

ContrastgOrig6 <- sd.y*Contrastg6
credLowerOrig6 <- sd.y*credLower6 
credUpperOrig6<- sd.y*credUpper6

par(mfrow = c(1,1),mai = c(1.02,0.9,0.82,0.42))
plot(0,0,type = "n",bty = "l",xlim = range(xgOrig),
     range(c(credLowerOrig6,credUpperOrig6)),
     main="Black-Hispanic Contrast Curve Contrast Curve",
     xlab = "Age",
     ylab = " Spinal Bone Mineral Density",
     cex.lab = cex.labVal,cex.axis = cex.axisVal)

polygon(c(xgOrig,rev(xgOrig)),c(credLowerOrig6,rev(credUpperOrig6)),
        col = varBandCol,border = FALSE)

lines(xgOrig,ContrastgOrig6,lwd = 2,col = estFunCol)
abline(0,0,col = "slateblue",lwd = 2)
```

### c
```{r}
library(mgcv)
fit <- gamm(spnbmd~black+hispanic+white+s(age),random = list(idnum = ~1),data=femSBMD)
plot(fit$gam,shade = TRUE,shade.col = "palegreen",bty = "l")
summary(fit)
```

```{r}
col_names <- colnames(femSBMD)[5:8]
femSBMD$race <- apply(femSBMD[, 5:8], 1, function(x) {paste(col_names[x == 1], collapse = "-")})
femSBMD$race <- factor(femSBMD$race)
res<-residuals(fit$lme)
boxplot( res ~ race, data = femSBMD, main = "Residuals by Race", xlab = "Race", ylab = "Residuals")
```

It seems that the race which we chose as the pivot fits the best as per the residual, in this case it is asian race as it has the least amount of extremums. Hispanic is close. 

### d

As we can see in the gamm approach we pivot one group as the referance group but this is not done on the MCMC approach. THis is the difference in gamm fitting vs mcmc fitting mechanism. The disadvantage of the gamm and the advantage of MCMC is that it allows more symmetrical measures the random effects for the group. 

One advantage of gamm and the disadvantage of mcmc is that mcmc takes posterior distribution and samples from it which makes it computationally very time consuming whilen gam is moderately fast in comparison.