---
title: "HW7"
author: "ASG"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE, cache=TRUE}
knitr::opts_chunk$set(echo = TRUE)
```
## 1

The very first thing one should look into when selecting the model terms , is the domain knowledge of the data , the hypothesis problem and the dataset itself. It is very important to first weed out the terms based on the domain knowledge of the data and the hypothesis one is working on. Rest steps can follow after this necessary first step.

## 3.2

### a
```{r}
set.seed(1) ; n <- 500 ; error <- rnorm(n,0,0.5)
x1 <- runif(n) ; x2 <- runif(n) ; x3 <- runif(n)
x4 <- runif(n) ; x5 <- runif(n) ; x6 <- runif(n)
x7 <- runif(n) ; x8 <- runif(n) ; x9 <- runif(n)
f4 <- function(x) return(x + dnorm(x,0.5,0.25))
f5 <- function(x) return(x + 0.5*dnorm(x,0.5,0.25))
f6 <- function(x) return(x + 0.1*dnorm(x,0.5,0.25))
y <- x1 + x2 + x3 + f4(x4) + f5(x5) + f6(x6) + error
```

### b
```{r}
library(gam)
gamObj <- gam(y ~ x1 + x2 + x3 + x4 + x5 + x6 + x7 + x8 + x9)
stepFit <- step.Gam(gamObj,scope =list("x1" = ~ 1 + x1 + s(x1,df = 2),"x2" = ~ 1 + x2 + s(x2,df = 2),"x3" = ~ 1 + x3 + s(x3,df = 2),"x4" = ~ 1 + x4 + s(x4,df = 2),"x5" = ~ 1 + x5 + s(x5,df = 2),"x6" = ~ 1 + x6 + s(x6,df = 2),"x7" = ~ 1 + x7 + s(x7,df = 2),"x8" = ~ 1 + x8 + s(x8,df = 2),"x9" = ~ 1 + x9 + s(x9,df = 2)))

```

### c
```{r}
print(names(stepFit$"model")[-1])

correct_fxn<-function(fnames){
  correct<-c(
  "x1" %in% fnames,
  "x2" %in% fnames,
  "x3" %in% fnames,
  "s(x4, df = 2)" %in% fnames,
  "s(x5, df = 2)" %in% fnames,
  "s(x6, df = 2)" %in% fnames,
  !("s7" %in% fnames | "s(x7, df = 2)" %in% fnames),
  !("s8" %in% fnames | "s(x8, df = 2)" %in% fnames),
  !("s9" %in% fnames | "s(x9, df = 2)" %in% fnames)
  )
  as.numeric(correct)
}
ans<-correct_fxn(names(stepFit$"model")[-1])
print(ans)
```


### d
```{r}
library(gam)
prop<-rep(0,9)
for (i in 1:100){
  set.seed(i) ; n <- 500 ; error <- rnorm(n,0,0.5)
  x1 <- runif(n) ; x2 <- runif(n) ; x3 <- runif(n)
  x4 <- runif(n) ; x5 <- runif(n) ; x6 <- runif(n)
  x7 <- runif(n) ; x8 <- runif(n) ; x9 <- runif(n)
  f4 <- function(x) return(x + dnorm(x,0.5,0.25))
  f5 <- function(x) return(x + 0.5*dnorm(x,0.5,0.25))
  f6 <- function(x) return(x + 0.1*dnorm(x,0.5,0.25))
  y <- x1 + x2 + x3 + f4(x4) + f5(x5) + f6(x6) + error
  gamObj <- gam(y ~ x1 + x2 + x3 + x4 + x5 + x6 + x7 + x8 + x9)
  stepFit <- step.Gam(gamObj,trace=FALSE,scope =list("x1" = ~ 1 + x1 + s(x1,df = 2),"x2" = ~ 1 + x2 + s(x2,df = 2),"x3" = ~ 1 + x3 + s(x3,df = 2),"x4" = ~ 1 + x4 + s(x4,df = 2),"x5" = ~ 1 + x5 + s(x5,df = 2),"x6" = ~ 1 + x6 + s(x6,df = 2),"x7" = ~ 1 + x7 + s(x7,df = 2),"x8" = ~ 1 + x8 + s(x8,df = 2),"x9" = ~ 1 + x9 + s(x9,df = 2)))
  vars<-correct_fxn(names(stepFit$"model")[-1])
  prop=prop+vars
}
print(prop/100)
detach(package:gam)
```

### e
```{r}
library(gam)
prop2<-rep(0,9)
for (i in 1:100){
  set.seed(i) ; n <- 500 ; error <- rnorm(n,0,1)
  x1 <- runif(n) ; x2 <- runif(n) ; x3 <- runif(n)
  x4 <- runif(n) ; x5 <- runif(n) ; x6 <- runif(n)
  x7 <- runif(n) ; x8 <- runif(n) ; x9 <- runif(n)
  f4 <- function(x) return(x + dnorm(x,0.5,0.25))
  f5 <- function(x) return(x + 0.5*dnorm(x,0.5,0.25))
  f6 <- function(x) return(x + 0.1*dnorm(x,0.5,0.25))
  y <- x1 + x2 + x3 + f4(x4) + f5(x5) + f6(x6) + error
  gamObj <- gam(y ~ x1 + x2 + x3 + x4 + x5 + x6 + x7 + x8 + x9)
  stepFit <- step.Gam(gamObj,trace=FALSE,scope =list("x1" = ~ 1 + x1 + s(x1,df = 2),"x2" = ~ 1 + x2 + s(x2,df = 2),"x3" = ~ 1 + x3 + s(x3,df = 2),"x4" = ~ 1 + x4 + s(x4,df = 2),"x5" = ~ 1 + x5 + s(x5,df = 2),"x6" = ~ 1 + x6 + s(x6,df = 2),"x7" = ~ 1 + x7 + s(x7,df = 2),"x8" = ~ 1 + x8 + s(x8,df = 2),"x9" = ~ 1 + x9 + s(x9,df = 2)))
  vars<-correct_fxn(names(stepFit$"model")[-1])
  prop2=prop2+vars
}
print(prop2/100)
detach(package:gam)
```

### f
```{r}
y2<-prop2/100
y1<-prop/100
x<-seq(1,9)
plot(y2,col=2,xlab="Coefficient terms",ylab="Probability")
points(y1,col=3,pch=3)
legend("bottomright", legend=c("Simulation 1", "Simulation 2"),
       col=c("red", "green"), pch=c(1,3), cex=0.8)
```

## 3.4

### a
```{r}
library(aplore3) ; data(icu) ; help(icu)
str(icu)
```

### b
```{r}
library(gam)
fitInitial <- gam:::gam(sta ~age+gender+race+ser+can+crn+inf+cpr+sys+hra+pre+type+fra+po2+ph+pco+bic+cre+loc,family=binomial,data = icu)
stepFit <- step.Gam(fitInitial, scope =
                      list(
                      "age" = ~ 1 + age + s(age,2), 
                      "gender" = ~ 1 + gender,
                      "race" = ~ 1 + race , 
                      "ser" = ~ 1 + ser, 
                      "can" = ~ 1 + can, 
                      "crn" = ~ 1 + crn, 
                      "inf" = ~ 1+ inf,
                      "cpr" = ~ 1 + cpr,
                      "sys" = ~ 1 + sys + s(sys,2),
                      "hra" = ~ 1+ hra + s(hra,2),
                      "pre" = ~ 1 + pre,
                      "type" = ~ 1 + type ,
                      "fra" = ~ 1 + fra,
                      "po2" = ~ 1 + po2, 
                      "ph" = ~ 1 + ph,
                      "pco" = ~ 1 + pco,
                      "bic" = ~ 1 + bic,
                      "cre" = ~ 1 + cre ,
                      "loc" = ~ 1 + loc
                      ))
print(names(stepFit$"model")[-1])
detach(package:gam)
```

### c
```{r}
library(mgcv)
fitgam<-mgcv:::gam(sta~s(age)+as.factor(gender)+as.factor(can)+as.factor(cpr)+sys+as.factor(pre)+as.factor(type)+as.factor(ph)+as.factor(pco)+as.factor(loc),family=binomial,data=icu)
summary(fitgam)
gam.check(fitgam)

```

### d
```{r}
newdatadf<-data.frame(
  gender="Female",
  age=79,
  race="White",
  ser="Medical",
  can="No",
  crn="No",
  inf="No",
  cpr="No",
  sys=228,
  hra=94,
  pre="No",
  type="Emergency",
  fra="No",
  po2="> 45",
  ph=">= 7.25",
  pco="> 45",
  bic="< 18",
  cre="> 2.0",
  loc="Coma"
)
predObjdir <- predict(fitgam,newdata = newdatadf,
                      type = "response",se.fit = TRUE)
print(predObjdir)
```
## 3.7

### a
```{r}
library(kernlab) ; data(spam) ; help(spam)
print(names(spam))
```

### b
```{r}
set.seed(1) ; nTest <- 1000
indsTest <- sample(1:nrow(spam),nTest,replace = FALSE)
indsTrain <- setdiff(1:nrow(spam),indsTest)
spamTest <- spam[indsTest,]
spamTrain <- spam[indsTrain,]
```

### c
```{r}
fitTrainFullGLM <- glm(type ~ .,family = binomial,data = spamTrain)
print(summary(fitTrainFullGLM))
```

### d
```{r}
library(gam)
fitgam3<-gam:::gam(type ~ .,family = binomial,data = spamTrain)
spam_scope<-gam.scope(spamTrain,arg="df=2",response=58)
stepfit3<-step.Gam(fitgam3,scope=spam_scope)
detach(package:gam)
```

### d
```{r}
func<-names(stepfit3$"model")[-1]
print(func)
library(mgcv)
fitmgcvgam3<-mgcv:::gam(type~make+address+all+num3d+s(our)+over+s(remove)+internet+order+mail+receive+will+people+report+addresses+free+business+email+you+credit+your+font+num000+s(money)+s(hp)+s(george)+hpl+s(num650)+lab+labs+num857+data+num415+num85+technology+num1999+parts+pm+direct+cs+meeting+original+project+re+edu+table+conference+charSemicolon+charRoundbracket+charSquarebracket+s(charExclamation)+s(charDollar)+charHash+capitalAve+capitalLong+s(capitalTotal),family=binomial,data=spamTrain)
summary(fitmgcvgam3)
```

### e
```{r}
library(caret)
pdata <- predict(fitmgcvgam3, newdata = spamTest, type = "response")
SpamTestpred = rep("nonspam", dim(spamTest)[1])
SpamTestpred[pdata>0.5]="spam"
p<-confusionMatrix(data = as.factor(SpamTestpred), reference = spamTest$type)
print(p)
print("classification error:")
print(1-p$overall[1])
```

### f
```{r}
library(mgcv)
fitmgcvgam4<-mgcv:::gam(type~s(make)+s(address)+s(all)+s(num3d)+s(our)+s(over)+s(remove)+s(internet)+s(order)+s(mail)+s(receive)+s(will)+s(people)+s(report)+s(addresses)+s(free)+s(business)+s(email)+s(you)+s(credit)+s(your)+s(font)+s(num000)+s(money)+s(hp)+s(george)+s(hpl)+s(num650)+s(lab)+s(labs)+s(num857)+s(data)+s(num415)+s(num85)+s(technology)+s(num1999)+s(parts)+s(pm)+s(direct)+s(cs)+s(meeting)+s(original)+s(project)+s(re)+s(edu)+s(table)+s(conference)+s(charSemicolon)+s(charRoundbracket)+s(charSquarebracket)+s(charExclamation)+s(charDollar)+s(charHash)+s(capitalAve)+s(capitalLong)+s(capitalTotal),family=binomial,data=spamTrain)
```
```{r}
library(caret)
pdata <- predict(fitmgcvgam4, newdata = spamTest, type = "response")
SpamTestpred = rep("nonspam", dim(spamTest)[1])
SpamTestpred[pdata>0.5]="spam"
p<-confusionMatrix(data = as.factor(SpamTestpred), reference = spamTest$type)
print(p)
print("classification error:")
print(1-p$overall[1])
```
### g
For this scenario, One rule is not significantly better than other. This is seen because the accuracy/classification error seems to be very much similar. So adding spline coefficients for all the terms didnt significantly improve the model but it surely increased the computational time. One more reason maybe because of the overly complicated model with lot of features, doesnt really affect when all the terms are added for spline coefficients 